---
title: "The Stepping Habits of Anonymous"
author: "kb"
date: "November 27, 2016"
output: html_document
---

This report analyzes data from Anonymous's walking habits.

First, we'll need to load the data from our working directory:

```{r}
setwd("/home/thea/Desktop/R working directory/ReproResearch/Wk2")
ourfile <- read.csv("activity.csv", header=TRUE)
```

# 1) What is the mean total number of steps taken per day?

Our first question is how many total steps Anonymous walks per day. What are the mean and median steps per day?

We can generate three lists to show us this information:

```{r}
sumsteps <- aggregate(ourfile$steps~ourfile$date, ourfile, sum)
meansteps <- aggregate(ourfile$steps~ourfile$date, ourfile, mean)
mediansteps <- aggregate(ourfile$steps~ourfile$date, ourfile, median)

colnames(sumsteps) <- c("Date", "TotalSteps")
meansteps[,2] <- as.integer(meansteps[,2])

stepswnas <- data.frame(sumsteps,meansteps[,2],mediansteps[,2])
colnames(stepswnas) <- c("Date", "Total Steps", "Average Steps", 
                         "Median Steps")
```

Here are the first 20 rows of data compiled into one table. Notice rows with missing data have been omitted:

```{r stepswnas, echo=FALSE}
head(stepswnas,20)
```

Here are the Total Steps per day depicted in a histogram:

```{r}
barplot(sumsteps[,2], names.arg=sumsteps[,1],
        main="Total Steps per Day, excluding NAs",
        xlab="Date", ylab="Steps")
```

# 2) What is the average daily activity pattern?

Next, let's study the average number of steps taken per 5-minute interval. This will show at which times of day Anonymous is most active, on average.

```{r}
stepsperint <- aggregate(ourfile$steps~ourfile$interval, ourfile, mean)
colnames(stepsperint) <- c("interval","AvgSteps")
```

Here is a time series plot of when Anonymous is most active:

```{r}
plot(stepsperint$interval, stepsperint$AvgSteps, type="l", 
     xlab="5-min interval", ylab="Steps taken", 
     main="Time lapse of average steps over 24 hours")
```

At which time of day is Anonymous most active, on average?

```{r}
maxstep <- which(stepsperint[,2] == max(stepsperint[,2])) #placeholder of max steps
int <- stepsperint[maxstep,1]
timehr <- paste(substr(int,1,1),":",substr(int,2,3)," am", sep="")
```

Thus, Anonymous takes the most steps, on average, around **`r timehr`**.

# 3) Inputing missing values

Now we'd like to address missing data points. First, let's identify all missing values in our data.

```{r}
nas <- ourfile[is.na(ourfile$steps),] #data frame with all na values
totalnas <- length(nas[,1])
percentnas <- round(totalnas/length(ourfile[,1])*100,2)
```

It looks like there are `r totalnas` missing rows in our data, or `r percentnas`% of all the rows. We can fill these in with the average steps per 5-min interval that Anonymous usually walks on other days.

```{r}
#1 create big vector w "answer" column adjacent to Steps + reorder
answerkey <- merge(ourfile,stepsperint)
answerkey <- answerkey[with(answerkey, order(date,interval)),]

#2 if steps = na, replace w answerkey
fullsteps <- with(answerkey,ifelse(is.na(steps),AvgSteps,steps))

#3 stitch "fullsteps" back into full data frame
complete <- data.frame(answerkey[,3],answerkey[,1],fullsteps)
complete[,3] <- as.integer(complete[,3])
colnames(complete) <- c("Date", "Interval", "Steps")
```

The results is a "filled in" data set, where N/A values have been replaced by the average value for that interval. Here is a sample of the top 20 rows:

```{r complete, echo=FALSE}
head(complete,20)
```

Now, let's calculate the Total, Average, and Median steps taken per day of this complete data set:

```{r}
sumfullsteps <- aggregate(complete$Steps~complete$Date, complete, sum)
meanfullsteps <- aggregate(complete$Steps~complete$Date, complete, mean)
medianfullsteps <- aggregate(complete$Steps~complete$Date, complete, median)

sumfullsteps[,2] <- sapply(sumfullsteps[,2],as.integer)
meanfullsteps[,2] <- sapply(meanfullsteps[,2],as.integer)
medianfullsteps[,2] <- sapply(medianfullsteps[,2],as.integer)

colnames(sumfullsteps) <- c("Date","TotSteps")
colnames(meanfullsteps) <- c("Date", "AvgSteps")
colnames(medianfullsteps) <- c("Date","MedSteps")
```

We can compare the "adjusted" data to the original data set with NAs.

```{r}
#1 clean data
colnames(stepswnas) <- c("Date", "TotStepsNA", "AvgStepsNA", "MedStepsNA")
stepswnas[,3] <- as.integer(stepswnas[,3])

#2 merge all data frames
merge1 <- merge(stepswnas, sumfullsteps, all=T)
merge2 <- merge(meanfullsteps, medianfullsteps)
allstats <- merge(merge1,merge2, all=T)

#3 reorder columns for clarity
allstats <- allstats[,c(1,2,5,3,6,4,7)]
```

Which gives us the following data frame, where headers including 'NA' are the original, unadjusted data. Here is an excerpt of the top 20 rows:

```{r allstats, echo=F}
head(allstats,20)
```

```{r, include=F}
stepsthrustudyNA <- sum(allstats$TotStepsNA, na.rm=T)
stepsthrustudy <- sum(allstats$TotSteps)
avgstepsthrustudyNA <- as.integer(stepsthrustudyNA/length(allstats[,1]))
avgstepsthrustudy <- as.integer(stepsthrustudy/length(allstats[,1]))
```

Notice the Total Steps taken across the course of the study increase from `r stepsthrustudyNA` with NAs included, to `r stepsthrustudy` with NA values estimated from remaining data. This amounts to an average of `r avgstepsthrustudyNA` steps per day with NAs included, and `r avgstepsthrustudy` steps per day with "filled in" data.

Finally, let's examine a histogram of the Total Steps per day.

```{r}
with(sumfullsteps,{
    barplot(TotSteps, names.arg=Date, 
            col=adjustcolor("blue",alpha.f=.5),
            main="Total Steps per Day",
            xlab="Date", ylab="Steps")
})
```

# 4) Are there differences in activity patterns between weekdays and weekends?

First, let's add a column to our "filled in" data frame indicating whether each date is a weekday or weekend.

```{r}
#1 column for which day of week each date is
complete[,4] <- sapply(as.Date(complete[,1]),weekdays)
colnames(complete)[4] <- "DayofWeek"
complete[,3] <- as.integer(complete[,3])

#2 column: weekday or weekend?
complete[,5] <- ifelse(complete[,4]=="Saturday" | 
                       complete[,4]=="Sunday", "weekend", 
                       "weekday")
colnames(complete)[5] <- "DayType"
```

Now we can prepare our data for visualization by aggregating it by weekday vs weekend and 5-min interval:

```{r}
wklysteps <- aggregate(complete$Steps~complete$Interval+DayType, complete, mean)
wklysteps[,3] <- as.integer(wklysteps[,3])
colnames(wklysteps) <- c("Interval", "DayType", "Steps")
```

Here is the code to plot a time series of when Anonymous is most active, segregated by weekday vs weekend:

```{r}
library(ggplot2)
wkdayplot <- 
    with(wklysteps,{
        qplot(Interval, Steps, facets = DayType~., geom="line",
              main="Average Steps Taken by Weekday/Weekend")
    })
```

And the plot itself:

```{r wkdayplot, echo=F}
wkdayplot
```

# THE END

